---
title: 'Lecture 3: <br> Manipulating data; lists; functions; if-else statements'
author: "Dr. Osita Onyejekwe"
date: "Spring 2021"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
    highlight: Rpub
---

## Agenda

- Wrap up of Lecture 2 content  # done 
- More on data frames
- Basic tidyverse (dplyr) commands
- Lists
- Writing functions in R
- If-else statements
- R coding style

## Wrapping up Lecture 2 content  


## Load tidyverse
- Most of the functions we're using just come from `dplyr`, but we'll load all of `tidyverse` anyway
```{r, message=FALSE, warning = FALSE}
#install.packages('tidyverse')
library(tidyverse)
```


## Grab a toy dataset from MASS library
- Rather than loading the full MASS library, we'll use the `::` syntax to pull a specific object/function from the library

- Loading all of MASS with `library(MASS)` after tidyverse is loaded has the unintended consequence of replacing the `dplyr` select command with the `MASS` select command.  This is BAD, and leads to errors.

```{r, message=FALSE, warning = FALSE}
Cars93 <- MASS::Cars93
head(Cars93, 3)

```

## Adding a column: `mutate()` function from `dplyr`

- **`mutate()`** returns a new data frame with columns modified or added as specified by the function call

```{r}
Cars93.metric <- mutate(Cars93,
                        KMPL.city = 0.425 * MPG.city, KMPL.highway = 0.425* MPG.highway)

tail(names(Cars93.metric))

```

- Our data frame has two new columns, giving the fuel consumption in km/l

## Another approach
```{r}
# Add a new column called KMPL.city.2

Cars93.metric$KMPL.city.2<- 0.425* Cars93$MPG.city

tail(names(Cars93.metric))

```

- Let's check that both approaches did the same thing

```{r}

identical( Cars93.metric$KMPL.city , Cars93.metric$KMPL.city.2)
identical(5,4)


```


## Changing levels of a factor: `recode()`
```{r}
manufacturer <- Cars93$Manufacturer
head(manufacturer, 8)

```


We'll use the **`recode()`** function from the `dplyr` library, which gets loaded when you load `tidyverse`.

```{r}
# Map Chevrolet, Pontiac and Buick to GM


manufacturer.combined <- recode(manufacturer, "Chevrolet" = "GM", "Pontiac"= "GM", "Buick"= "GM")
head(manufacturer.combined, 8)

# Stopped here on 02/03/2021

```



## Another example: `recode_factor()`
- A lot of data comes with integer encodings of levels

- You may want to convert the integers to more meaningful values for the purpose of your analysis

- Let's pretend that in the class survey 'Program' was coded as an integer with 1 = MISM, 2 = Other, 3 = PPM

```{r}
# practice on your own with numeric values for the categories for Program 
# Load data survey
# Recode Program to have integer codings 
survey<- read.csv('/Users/osita/OneDrive/Desktop/STAT 2600 SPRING 2021/Coding/survey_data2020.csv')
survey<- mutate(survey, Program=as.numeric(Program))
head(survey)


```

## Example continued: recode_factor()

- Here's how we would get back the program codings using **`recode_factor()`**, a variant of `recode` that returns a factor, with elements ordered according to the mapping order.  
- Note the backticks ` ` around the numbers, which are necessary for parsing

```{r}
survey<- mutate(survey,  Program = recode_factor(Program, 
                                                 `3` = "PPM",
                                                 `1` = "MISM",
                                                 `2` = "Other"))
head(survey)
# practice on your own 

```


## Some more data frame summaries: `table()` function

- Let's revisit the Cars93 dataset

- The **`table()`** function builds **contingency tables** (i.e., count tables) showing counts at each combination of factor levels

```{r}
table(Cars93$AirBags)

```
## 
```{r}
table(Cars93$Origin)

table(Cars93$AirBags, Cars93$Origin)

```

- Looks like US and non-US cars had about the same distribution of AirBag types

- Later in the class we'll learn how to do a hypothesis tests on this kind of data


## Alternative syntax

- When `table()` is supplied a data frame, it produces contingency tables for all combinations of factors 

```{r}
#head(Cars93[c("AirBags", "Origin")], 3)
table(Cars93[c("AirBags", "Origin")])

```

## Tidy count tables: `count()`

If we're going to be plotting or further analysing our results, it is helpful to have them in a data frame instead of a tabular layout.  That's where the **`count()`** function comes in.

```{r}
Cars93 %>% count(AirBags)
Cars93 %>% count(AirBags, Origin)

```

## Basics of lists

> A list is a **data structure** that can be used to store **different kinds** of data

- Recall: a vector is a data structure for storing *similar kinds of data*

- To better understand the difference, consider the following example.

```{r}

my.vector.1 <- c("Michael", 165, TRUE)  # (name, weight, is.male)
my.vector.1

typeof(my.vector.1)  #All the elements ae not character strings: 

```

## Lists vs. vectors

```{r}
my.vector.2 <- c(FALSE, TRUE, 27)  #( is.male, is.citizen, age)
typeof(my.vector.2)

```

- Vectors expect elements to be all of the same type (e.g., `Boolean`, `numeric`, `character`)

- When data of different types are put into a vector, the R converts everything to a common type

## Lists

- To store data of different types in the same object, we use lists

- Simple way to construct lists: use **`list()`** function

- (We'll learn about functions like `map` and `map_chr` soon.)

```{r}

my.list<- list("Michael", 165, TRUE)
my.list

map_chr(my.list, typeof)

```

## Named elements
```{r}
patient.1<- list(name= "Michael", weight = 165, is.male=TRUE)
patient.1

```

## Referencing elements of a list (similar to data frames)
```{r}
patient.1$name  # Get "name" element (returns a string)
patient.1[["name"]]   # get "name" element (returns a string)
patient.1["name"] # Get "name" slice (returns a list)
 c(typeof(patient.1$name), typeof(patient.1["name"]))
 # we stopped here on 02/05/2021
```

## Functions
- We have used a lot of built-in functions: `mean()`, `subset()`, `plot()`, `read.table()`...

- An important part of programming and data analysis is to write custom functions

- Functions help make code **modular**

- Functions make debugging easier

- Remember: this entire class is about applying *functions* to *data*

## What is a function?

> A function is a machine that turns **input objects** (arguments) into an **output object** (return value) according to a definite rule.

- Let's look at a really simple function

```{r}
addOne <- function(x) {x+1}

```

- `x` is the **argument** or **input**

- The function **output** is the input `x` incremented by 1

```{r}
addOne(12)
```

## More interesting example

- Here's a function that returns a % given a numerator, denominator, and desired number of decimal values

```{r}
calculatePercentage <- function(x,y,d){decimal <- x/y    # calculate decimal value 
round(100* decimal, d)}     # convert to % and round to d digits
calculatePercentage(27,80,1)

```

- If you're calculating several %'s for your report, you should use this kind of function instead of repeatedly copying and pasting code

## Function returning a list
- Here's a function that takes a person's full name (FirstName LastName), weight in lb and height in inches and converts it into a list with the person's first name, person's last name, weight in kg, height in m, and BMI.

```{r}
createPatientRecord<- function(full.name, weight, height){
name.list <- strsplit(full.name, split = " ")[[1]]
first.name <- name.list[1]
last.name<- name.list[2]
weight.in.kg <- weight / 2.2
height.in.m <- height * 0.0254
bmi <- weight.in.kg/(height.in.m  ^ 2)
list(first.name= first.name, last.name= last.name, weight=weight.in.kg, height= height.in.m, bmi=bmi)
}

```
 
## Trying out the function
```{r}
createPatientRecord("Michael Smith", 185, 12*6+1)
```

## Another example: 3 number summary
- Calculate mean, median and standard deviation
```{r}
threeNumberSummary<- function(x) { 
  c(mean = mean(x), median=median(x), sd=sd(x))
  }
x<- rnorm(100, mean = 5, sd=2) # Vector of 100 normals with mean  =5 , sd = 2 
threeNumberSummary(x)
```


## If-else statements

- Oftentimes we want our code to have different effects depending on the features of the input

- Example: Calculating a student's letter grade
  - If grade >= 90, assign A
  - Otherwise, if grade >= 80, assign B
  - Otherwise, if grade >= 70, assign C
  - In all other cases, assign F

- To code this up, we use if-else statements

## If-else Example: Letter grades
```{r}


calculateLetterGrade <- function(x) { 
  
  if(x >= 90)  {
grade <-  "A"
} else if (x >= 80) { 
  
grade <- "B"

} else if (x >= 70) { 
  grade <- "C"

} else {
  
  grade <- "F"
}
  grade 
}

course.grades <- c(92, 78, 87, 91, 62)
map_chr(course.grades, calculateLetterGrade)

```

## `return()`

- In the previous examples we specified the output simply by writing the output variable as the last line of the function

- More explicitly, we can use the **`return()`** function

```{r}
addOne<- function(x) { 
  return(x+1)
  
  }
addOne(12)


```

- We will generally avoid the `return()` function, but you can use it if necessary or if it makes writing a particular function easier.
- Google's style guide suggests explicit returns.  Most do not. 

## Reminders

- Homework 1 **due 11:59PM MST on Wednesday FEB 10 **

- Lab 3 is due Friday 11:59PM Feb 05

- If you have questions, feel free to post on the Discord Channel or attend office hours