\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=2cm]{geometry}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{mathtools}
\usepackage{MnSymbol}
\usepackage{pgfplots}
\usepackage{physics}
\usepackage{titling}

\pgfplotsset{compat=1.6}

\begin{document}
%--------------------------------------title--------------------------------------
\begin{flushright} STAT 2600 Recitation\\* 9/22/2020 \end{flushright}
\begin{center}
\textbf{Recitation Week 5 Problems}
\end{center}
%--------------------------------------body--------------------------------------
 Today imagine you are a Premiere League soccer coach and you want to know which players to consider buying / trading for next season. You will use data analysis skills you have learned to provide support for your argument. To begin, import the data from the player\_data.csv file.
 
    \begin{enumerate}

	\item[1) ]{*Skip this if it takes too much time* Clean the data in each variable (you may want to use the substr() function) - remove any unnecessary spaces, replace any non-standard characters with an underscore "\_".}
	
	\item[2) ]{(a) Which player had the most total points over the season? (hint: check out the arrange() function). (b) Who had the highest points per minute? (hint: you will need to calculate points per minute).}
	
	\item[3) ]{Which team had the most points over the season? To find this, you will need to (a) import the team\_lookup.csv file and merge it onto your data, then (b) summarize or "aggregate" your data to the team level (as opposed to the player level).}
	
	\item[4) ]{*Skip if you are running out of time* Come up with a metric to measure players' performances controlling for what team they were on. Who scored best? (e.g. you may want to calculate who scored highest above their team average...).}
	
	\item[5) ]{Which players would you draft / purchase? Are these the most expensive players? If not why do you think they may be undervalued?}
   
     \end{enumerate}
\end{document}